---
title: "Class 11: Halloween Mini-Project"
author: "Emily Rodriguez"
format: pdf
---



```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
candy$fruity
```

```{r}
sum(candy$fruity)
```

> Q. What are these fruity candy?

We can use the == 
```{r}
rownames( candy[ candy$fruity == 1, ])
```


## How often does my favorite candy win?

```{r}
candy["Twix",]$winpercent
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Starburst",]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the `winpercent` column is on a 0:100 scale and all other appear to be 0:1 scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero here means that the candy is not classified as containing chocolate.

> Q8. Plot a histogram of winpercent values

In base R graphics:

```{r}
hist(candy$winpercent)
```


```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 10)
```


>Q9. Is the distribution of winpercent values symmetrical?

No, the distribution is not symmetrical.

>Q10. Is the center of the distribution above or below 50%?

The center of distribution is below 50% with a mean:

```{r}
mean(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

To answer this question I will need to:
- "subset" (a.k.a. "select", "filter") the candy dataset to just chocolate candy
- get their winpercent values
- calculate the mean of these.

Then do the same for fruity candy and compare. 

```{r}
# Filter/select/subset to just chocolate rows
chocolate.candy <- candy[as.logical(candy$chocolate),]

# Get their winpercent values
chocolate.winpercent <- chocolate.candy$winpercent

# Calculate their mean winpercent value
mean(chocolate.winpercent)
```

Now we will do the same steps for fruity candy.

```{r}
# Filter/select/subset to just fruity rows
fruity.candy <- candy[as.logical(candy$fruity),]

# Get their winpercent values
fruity.winpercent <- fruity.candy$winpercent

# Calculate their mean winpercent value
mean(fruity.winpercent)
```

On average, chocolate is higer ranked than fruity candy.


> Q12. Is this difference statistically significant?

```{r}
t.test(chocolate.winpercent, fruity.winpercent)
```

Yes, the difference is statistically significant.


## Overall Candy Rankings

There os a base R function called `sort()` for, guess what sorting vectors of input.

```{r}
x <- c(5, 2, 10)

#sort(x, decreasing = TRUE)
sort(x)
```

The buddy function to `sort()` that is often more useful is called `order()`. It returns the "indices" of the input that would result in it being sorted

```{r}
order(x)
x[ order(x) ]
```

>Q13. What are the five least liked candy types in this set?

I can order by `winpercent`

```{r}
ord <- order(candy$winpercent)
head( candy[ord, ], 5 )
```


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
ord <- order(candy$winpercent, decreasing = TRUE)
head( candy[ord, ], 5)
```


> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

>Q17. What is the worst ranked chocolate candy?

Sixlets 

> Q18. What is the best ranked fruity candy?

Starburst


## Taking a look at pricepercent

>Q. What is the best candy for the least money?

```{r}
my_cols[as.logical(candy$fruity)] = "red"
```


```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) 
```

Add some labels

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text()
```

To deal with overlapping labels I can use the **ggrepel** package.

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```



## 5 Exploring the correlation structure

Pearson correlation goea between -1 and +1 with zero indicating no correlation and the values close to one being very higly (anti) correlated.

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are anti-correlated.

>Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent ot chocolate and bar.


## 6 Principal Component Analysis

The base R function for PCA is called `prcomp()` and we can set "scale=TRUE/FALSE". 

```{r}
pca <-  prcomp(candy, scale=TRUE)
summary(pca)

```

The main result of PCA - i.e. the new PC plot (projection of candy on our new PC axis) is contained in `pca$x`

```{r}
pc <- as.data.frame(pca$x)

ggplot(pc) +
  aes(PC1, PC2, label=rownames(pc)) +
  geom_point(col=my_cols) +
  geom_text_repel(max.overlaps = 5)
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```


>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard and pluribus are picked up strongly by PC1 in the positive direction.

