---
title: "Class 8: Mini Project"
author: "Emily Rodriguez"
format: pdf
---

# Breast Cancer Project

Today we are going to explore some data from the University of Wisconsin Cancer Center on Breast biopsy data.

```{r}
wisc.data <- read.csv("WisconsinCancer.csv", row.names = 1)
head(wisc.data)
```

> Q. How many patient samples are in this dataset?

```{r}
nrow(wisc.data)
```

There are `r nrow(wisc.data)` patients in this dataset.

> Q. How many cancer (M) and non cancer (B) samples are there?

```{r}
table(wisc.data$diagnosis)
```

Save the diagnosis for later use as a reference to compare how well we do with PCA etc.

```{r}
diagnosis <- as.factor(wisc.data$diagnosis)
#diagnosis
```

Now exclude the diagnosis column from the data.

```{r}
wisc <- wisc.data[,-1]
```

> Q. How many "dimensions", "variables", "columns", are in this dataset?

```{r}
ncol(wisc)
```

# Principal Component Analysis (PCA)

To perform PCA in R we can use the `prcomp()` function. It takes as input a numeric dataset and optional `scale=FALSE/TRUE` argument.

We generally always wanst to set `scale=TRUE` but let's make sure by checking if the mean and standard deviation values are different across these 30 columns.

```{r}
round(colMeans(wisc))
```

```{r}
pca <- prcomp(wisc, scale=TRUE)
summary(pca)
```

```{r}
attributes(pca)
```

```{r}
plot(pca$x[,1],pca$x[,2], col=diagnosis)
```

```{r}
library(ggplot2)

x <- as.data.frame(pca$x)

ggplot(x) +
  aes(PC1, PC2, col = diagnosis) +
  geom_point()
```

> Q. How much variance is captured in the top 3 PCs?

They capture 76% if the total variance.

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr\$rotation\[,1\]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC.

```{r}
pca$rotation["concave.points_mean",1]
```

```{r}
attributes(pca)
```

# Combine PCA results with clustering.

We can use our new pCA variables (i.e. the scores along the PCs contained in t `pca$x`) as input for other methods such as clustering.

```{r}
# Hclust needs a distance matrix as input
d <- dist(pca$x[,1:3])

hc <- hclust(d, method= "ward.D2")
plot(hc)
```

To get our cluster membership vector we can use the `cutree()` function and specify a height (`h`) or number of groups (`k`).

```{r}
grps <- cutree(hc, h=80)
table(grps)
```

I want to find out how many diagnosis "M" and "B" are in each grp?

```{r}
table(diagnosis)
```

```{r}
table(diagnosis, grps)
```

We can also plot our results using our clustering vector `grps`.

```{r}
plot(pca$x[,1], pca$x[,2], col = grps)
```

```{r}
library(ggplot2)

x <- as.data.frame(pca$x)

ggplot(x) +
  aes(PC1, PC2) +
  geom_point(col = grps)
```

> Q15. What is the specificity and sensitivity of our current results?

specificity:
```{r}
24/(24+333)
```

sensitivity:
```{r}
179/(179+33)
```


## Predictions

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(pca, newdata=new)
npc
```

```{r}
plot(pca$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

We should prioritize patient 2 for a follow up because the results look the most like the malignant results.

